Reagents: expressing and composing fine-grained concurrency

### 2011-11-07 paper/paper.tex
nearly done

Ahmdahl's law tells us that sequential bottlenecks fundamentally limit our
profit from parallelism~\cite{?}.  In practice, the effect is amplified by
another factor: interprocessor communication, often in the form of cache
coherence.  When one thread waits on another, the program pays the cost of
lost parallelism \emph{and} an extra cache miss.  The extra misses can easily
accumulate to yield parallel \emph{slowdown}, more than negating the benefits
of the remaining parallelism.

@@

Another answer is: communicate efficiently.  The past two
decades\footnote{Herlihy/Wing~\cite{?}, Mellor-Crummey/Scott~\cite{?}} have
produced a sizable collection of algorithms for synchronization, communication,
and shared storage which minimize the use of memory bandwidth and avoid
unnecessary waiting.  This research effort has led to industrial-strength
libraries---\texttt{java.util.concurrent} (JUC) the most prominent---offering a
wide range of concurrency primitives appropriate for fine-grained parallelism.

@@

Finally, it is worth noting in passing that reagents generalize both Concurrent
ML~\cite{?} and Transactional Events~\cite{?}, yielding the first lock-free
implementation for both.  We give a thorough discussion of this and other
related work---especially STM---in~\secref{related}.

@@

\figref{classic-treiber} gives a classic example of a fine-grained concurrent
data structure: Treiber's lock-free stack~\cite{?}.  The stack is represented as
an \emph{immutable} linked list.  Mutation occurs solely through the
\lstinline{head} pointer, represented here as an \lstinline{AtomicReference}.
The \lstinline{head} is mutated using \lstinline{compareAndSet} (here
abbreviated as \lstinline{cas}), which takes an expected value and a new value,
and atomically updates the reference if it has the expected value.  It returns
\lstinline{true} iff the update was successful.

@@

The \lstinline{push} and \lstinline{tryPop} operations are implemented in a
typical \emph{optimistic} style: they take a snapshot of the head, perform some
local computation with it, and then attempt to update it accordingly.  The
computation is optimistic because it is performed without holding a lock.  The
head might concurrently change, invalidating the \lstinline{cas} intended to
update it.  To account for possible interference, the snapshot-compute-update
code executes within a retry loop.  While the loop may retry forever, it can
only do so by repeatedly failing to \lstinline{cas}, which in turn means that an
unlimited number of other operations on the stack are succeeding.  In other
words, \lstinline{push} and \lstinline{tryPop} are formally \emph{lock
  free}~\cite{?}.

To successfully \lstinline{cas}, a core must acquire the relevant cache line in
\emph{exclusive} mode~\cite{?}, or something conceptually equivalent.  When
several cores attempt to \lstinline{cas} a common reference, they must
coordinate at the hardware level, using precious memory bandwidth and wasting
processor cycles.  A failed \lstinline{cas} is evidence that there is
\emph{contention} over a common location.  To increase the \lstinline{cas}
success rate and conserve memory bandwidth, fine-grained concurrent algorithms
employ a \emph{backoff} scheme, which here we have abstracted into a lightweight
\lstinline{Backoff} class.  The class encapsulates the simplest scheme:
busy-wait for a random amount of time that grows exponentially each time
\lstinline{once} is invoked.

@@

Both \lstinline{push} and \lstinline{tryPop} are \emph{total} operations: they
can in principle succeed no patter what state the stack is in, and fail (and
retry) only due to active interference from concurrent threads.  A true
\lstinline{pop} operation, on the other hand, is \emph{partial}: it is undefined
when the stack is empty.  Often this is taken to mean that the operation should
block until another thread moves the stack into a state on which the operation
is defined.  Partial operations introduce considerable complexity, because
\emph{all} of the operations on the data structure must potentially signal
blocked threads, depending on the changes being performed.  In some cases, it is
possible to cleverly treat signalling in essentially the same way as atomic
updates~\cite{?}.

Synchronization and signalling is also subject to cache and memory bandwidth
concerns.  For example, when a lock is released, how much memory traffic is
generated?  If the lock is implemented as a single bit which threads attempt to
\lstinline{cas} in a tight loop, any change to that bit will result in a cascade
of memory traffic---even though only one waiting thread can actually win and
acquire the lock.  The most important strategy for cutting down this traffic was
introduced by Mellor-Crummey and Scott~\cite{?}, who suggested that waiting
threads should place themselves in a lock-free queue, and the release operation
should  % TODO: room for this??

%talk about elimination backoff and blocking versions but don't show code.

%discuss the insight behind MCS locks

The goal of reagents is to systematize these techniques and others into a
uniform set of combinators that make algorithms easier to express, and allows
algorithms to be composed and tailored by users.  In the next section, we will
see how to write all of the above examples, concisely and composably.

@@

\begin{figure}
\begin{lstlisting}[mathescape=true,frame=single]
// combinators for shared state
upd:  Ref[A] => PartialFunction[(A,B), (A,C)] 
             => Reagent[B,C]
read: Ref[A] => Reagent[Unit, A]
cas:  Ref[A] * A * A => Reagent[Unit, Unit]

@@

// combinators for computation
ret: A => Reagent[Unit, A]
computed: PartialFunction[A, Reagent[Unit, B]]
          => Reagent[A,B]
postCommit: (A => Unit) => Reagent[A,A]
\end{lstlisting}
\nocaptionrule
\caption{The Reagent API}
\label{fig:combinators}
\end{figure}

Reagents are a new instance of an old idea: representing computations as data.
The computations being represented are fine-grained concurrent operations, so
a value of type \reagent{A}{B} represents a function from \lstinline{A} to
\lstinline{B} that internally interacts with a concurrent data structure
through mutation, synchronization, or both.

Unlike $\lambda$-abstractions, which encompass arbitrary code and treat that
code opaquely, reagents are built up from a limited set of \emph{combinators}
that are known to the implementation.  Existing reagents---for example, those
built by a concurrency expert---can be composed or tailored by library users
by applying additional combinators, without knowledge of the internal
implementation.  At any point a user can invoke a reagent, either as a
\emph{reactant} or a \emph{catalyst}---a distinction explained in
\secref{easy-examples}.

Reagents are patterned after Concurrent ML~\cite{?} and Haskell's STM
monad~\cite{?}, which represent (respectively) synchronous and atomic
operations as first-class values; we discuss the relationship in
\secref{related}.
%
Our contribution is giving a set of combinators appropriate for expressing and
composing fine-grained concurrent algorithms (\secref{easy-examples}) with a
clear semantics (\secref{semantics}), ... %TODO

@@

Memory is shared between reagents using the type \lstinline{Ref[A]} of
atomically-updatable references.\footnote{ These references are only
  accessible through reagent combinators, which allows them to provide
  functionality beyond the built-in \lstinline{AtomicReference[A]} type, such
  as the blocking logic we will see shortly.  }
The \lstinline{upd} combinator~(\figref{combinators}) represents simple atomic
updates to references.  It takes an \emph{update function}, which tells how to
transform a snapshot of the reference cell and some input into an updated
value for the cell and some output.  Using \lstinline{upd}, we can rewrite
\lstinline{TreiberStack} in a more readable and concise way:
\begin{lstlisting}
class TreiberStack[A] {
  private val head = new Ref[List[A]](Nil)
  val push: Reagent[A, Unit] = upd(head) {
    case (xs, x) => (x:xs, ())
  }
  val tryPop: Reagent[Unit, Option[A]] = upd(head) {
    case (x::xs, ()) => (xs, Some(x))
    case (Nil,   ()) => (Nil, None)
  }
}
\end{lstlisting}
In Scala, anonymous partial functions are written as a series of cases
enclosed in braces.  For \lstinline{push} and \lstinline{tryPop}, the case
analyses are exhaustive, so the update functions are in fact total.
\lstinline{Unit} is akin to \lstinline{void}: it is a type with a single
member, written \lstinline{()}.

Being reagents, \lstinline{push} and \lstinline{tryPop} are inert values.  They
can be invoked using the \lstinline{!} method provided by reagents, which is
read ``react''.  For a \lstinline{Reagent[A,B]} the \lstinline{!} method takes
an \lstinline{A} and returns a \lstinline{B}.  Scala permits infix notation for
methods, so we can use a \lstinline{TreiberStack} as follows:
\begin{lstlisting}
  var s = new TreiberStack[Int]; s.push ! 42
\end{lstlisting}
The key point is that when we invoke these reagents, we are executing
\emph{exactly} the same algorithms written by hand in \secref{background},
including the retry loop with exponential backoff; reagents systematize and
internalize common patterns of fine-grained concurrency.  But by exposing
\lstinline{push} and \lstinline{tryPop} as reagents, rather than methods, we
leave the door open for further composition and tailoring by a user of the
data structure (\secref{choice}, \secref{conjunction}).

While \lstinline{tryPop} handles both empty and nonempty stacks, we
can write a variant that drops the empty case:
\begin{lstlisting}
val pop: Reagent[Unit, A] = upd(head) {
  case (x::xs, ()) => (xs, x)
}
\end{lstlisting}
Now our update function is partial.  An invocation \lstinline{s.pop ! ()} will
block the calling thread unless or until the stack is nonempty.  Blocking and
signalling are entirely handled by the reagent implementation, using standard
techniques~(\secref{implementation}); there is therefore no risk of a lost
wakeup.  %TODO: add more here?

@@

Reagents interact through \emph{synchronous swap channels}, which
consist of two complementary \emph{endpoints}:
\begin{lstlisting}
  mkChan[A,B]: (Endpoint[A,B], Endpoint[B,A])
\end{lstlisting}
The combinator for communication is \lstinline{swap} (see
\figref{combinators}), which lifts an \lstinline{Endpoint[A,B]} to a
\lstinline{Reagent[A,B]}.  When two reagents communicate on opposite
endpoints, they provide messages of complementary type (\lstinline{A}
and \lstinline{B}, for example) and receive each other's value.  We
call a successful communication a \emph{reaction} between reagents.
On the other hand, if no complementary message is available,
\lstinline{swap} will block until a reaction can take place.  Once
again, the waiting and signalling protocol is managed internally by
the reagent implementation, so lost wakeups are impossible.

@@

If \lstinline{r} and \lstinline{s} are two reagents of the same type,
their \emph{choice} \lstinline{r + s} will behave like one of them,
nondeterministically, when invoked.  The most straightforward use of
choice is in the spirit of Unix's \lstinline{select}~\cite{?}: waiting
on several signals simultaneously, while consuming only one of them.
For example, if \lstinline{c} and \lstinline{d} are endpoints of the
same type, \lstinline{swap(c) + swap(d)} is a reagent that will accept
exactly one message, either from \lstinline{c} or from \lstinline{d}.
If neither endpoint has a message available, the reagent will block
until one of them does.  As usual, the blocking protocol is handled by
the reagent library.

A more novel use of choice is adding backoff strategies (of the kind
described in \secref{?}).  For example, we can build an elimination
backoff stack as follows:
\begin{lstlisting}
class EliminationStack[A] {
  private val s = new TreiberStack[A]
  private val (elimPop, elimPush) = mkChan[Unit,A]
  val push: Reagent[A,Unit] = s.push + swap(elimPush)
  val pop:  Reagent[Unit,A] = s.pop  + swap(elimPop)
}
\end{lstlisting}
Choice is left-biased,\footnote{The left bias makes choice more
  similar to \lstinline{orElse} in Haskell's STM than to CML's choice,
  but there are differences from both.  See \secref{related}.  } so
when \lstinline{push} is invoked, it will first attempt to push onto
the underlying Treiber stack.  If the underlying push fails (due to a
lost CAS race), \lstinline{push} will attempt to send a message along
\lstinline{elimPush}, \emph{i.e.}, to synchronize with a concurrent
popper.  If it succeeds, the \lstinline{push} reagent completes
without ever having modified its underlying stack.

@@

The double-pop reagent implicitly gives a solution to the Dining
Philosophers problem~\cite{?}: two resources are atomically acquired.
Internally, the reagent library ensures dead- and live-lock freedom
for conjoined reagents.

@@

In chemistry, a reagent is a participant in a reaction, and reagents are
further subdivided into \emph{reactants}, which are consumed during reaction,
and \emph{catalysts}, which enable reactions but are not consumed by them.

The \lstinline{!} operator invokes a reagent as a reactant: the invocation
lasts for a single reaction, and any messages sent by the reagent are
consumed.  But sometimes it is useful for reagents to persist beyond a single
reaction, \emph{i.e.}, to act as catalysts.  For example, the following
function creates a catalyst that merges input from two endpoints and sends the
resulting pairs to another endpoint:
\begin{lstlisting}
def zip(in1: Endpoint[Unit,  A], 
        in2: Endpoint[Unit,  B], 
        out: Endpoint[(A,B), Unit]) = 
  dissolve((swap(in1) ||| swap(in2)) >> swap(out))
\end{lstlisting}
The \lstinline{dissolve} function takes a \lstinline{Reagent[Unit, Unit]} and
introduces it as a catalyst.\footnote{For simplicity, we have not given a way
  to cancel catalysts after they have been introduced, but cancellation is
  easy to add.}  Operationally, in this example, that just means sending
messages along \lstinline{in1} and \listinline{in2} that are marked as
catalyzing messages, and hence are not consumed during reaction.  The upshot
is that senders along \lstinline{in1} will see the catalyzing messages, look
for messages along \lstinline{in2} to pair with, and ultimately send messages
along \lstinline{out} (and similarly in the other order).

@@

\elide{
class TreeBarrier {
  public readonly Synchronous.Channel[] Arrive;
  private readonly Join j; // create j, init chans ...
  public TreeBarrier(int n) {Wire(0, n-1, () => {});}
  private void Wire(int low, int high, Action Done) {
    if (low == high) j.When(Arrive[low]).Do(Done);
    else if (low + 1 == high)
      j.When(Arrive[low]).And(Arrive[high]).Do(Done);
    else { // low + 1 < high
      Synchronous.Channel Left, Right; // init chans       
      j.When(Left).And(Right).Do(Done);
      int mid = (low + high) / 2;
      Wire(low, mid, () => Left());
      Wire(mid + 1, high, () => Right()); 
    }
  }
}
}
\subsection{Dynamic reagents and \lstinline{cas}}
\label{sec:dynamic}

@@

\begin{figure}
\begin{lstlisting}[frame=single]
class MSQueue[A] {
  private case class Node(data: A, next: Ref[Node])
  private val head = new Ref(new Node(null))  // sentinel
  private val tail = new Ref(read(head) ! ()) // sentinel
  val tryDeq: Reagent[Unit, Option[A]] = upd(head) {
    case (Node(_, Ref(n@Node(x, _))), ()) => (n, Some(x))
    case (emp,                        ()) => (emp, None)
  }
  private def findAndEnq(n: Node): Reagent[Unit,Unit] = 
    read(tail) ! () match {
      case ov@Node(_, r@Ref(null)) => 
        cas(r, null, n) >>
        postCommit { _ => cas(tail, ov, n) !? () }
      case ov@Node(_, Ref(nv)) => 
        cas(tail, ov, nv) !? (); 
        findAndEnq(n)
    }
  val enq: Reagent[A, Unit] = computed { 
    (x: A) => findAndEnq(new Node(x, new Ref(null)))
  }
}
\end{lstlisting}
\nocaptionrule
\caption{The Michael-Scott queue, with reagents}
\label{fig:msqueue}
\end{figure}

@@

Here is a brief overview of the Michael-Scott queue.  The queue is represented
as a \emph{mutable} linked list, with a sentinel node at the head (front) of
the queue.  The head pointer always points to the current sentinel node; nodes
are dequeues by a CAS to this pointer, just like Treiber stacks (but lagged by
one node).  The tail of the queue is the unique node, reachable from the head
pointer, with a null \lstinline{next} pointer; thanks to the sentinel, such a
node is guaranteed to exist.  If the queue is empty, the same node will be the
head (sentinel) and tail.  Finally, as an optimization for enqueing, a tail
\emph{pointer} is maintained with the invariant that the real tail is always
reachable from it.  The tail pointer may lag behind, however, which allows the
algorithm to work using only single-word CAS instructions.

@@

As mentioned in~\secref{offers}, the initial continuation for reagents is the
\lstinline{Commit} continuation:

@@

\subsection{Methodology and experimental setup}

>>>>

Ahmdahl's law tells us that sequential bottlenecks fundamentally limit our
profit from parallelism\elide{~\cite{?}}.  In practice, the effect is amplified by
another factor: interprocessor communication, often in the form of cache
coherence.  When one thread waits on another, the program pays the cost of
lost parallelism \emph{and} an extra cache miss.  The extra misses can easily
accumulate to yield parallel \emph{slowdown}, more than negating the benefits
of the remaining parallelism.

@@

Another answer is: communicate efficiently.  The past two
decades\elide{\footnote{Herlihy/Wing~\cite{?}, Mellor-Crummey/Scott~\cite{?}}} have
produced a sizable collection of algorithms for synchronization, communication,
and shared storage which minimize the use of memory bandwidth and avoid
unnecessary waiting.  This research effort has led to industrial-strength
libraries---\texttt{java.util.concurrent} (JUC) the most prominent---offering a
wide range of primitives appropriate for fine-grained parallelism.

@@

Finally, it is worth noting in passing that reagents generalize both Concurrent
ML~\cite{Reppy1991} and Transactional Events~\cite{Donnelly2006}, yielding the
first lock-free implementation for both.  We give a thorough discussion of this
and other related work---especially STM---in~\secref{related}.

@@

\figref{classic-treiber} gives a classic example of a fine-grained concurrent
data structure: Treiber's lock-free stack~\cite{Treiber1986}.  The stack is
represented as an \emph{immutable} linked list.  Mutation occurs solely through
the \lstinline{head} pointer, represented here as an
\lstinline{AtomicReference}.  The \lstinline{head} is mutated using
\lstinline{compareAndSet} (here abbreviated as \lstinline{cas}), which takes an
expected value and a new value, and atomically updates the reference if it has
the expected value.  It returns \lstinline{true} iff the update was successful.

@@

The \lstinline{push} and \lstinline{tryPop} operations are implemented in a
typical \emph{optimistic} style: they take a snapshot of the head, perform some
local computation with it, and then attempt to update it accordingly.  The
computation is optimistic because it is performed without holding a lock.  The
head might concurrently change, invalidating the \lstinline{cas} intended to
update it.  To account for possible interference, the snapshot-compute-update
code executes within a retry loop.  While the loop may retry forever, it can
only do so by repeatedly failing to \lstinline{cas}, which in turn means that an
unlimited number of other operations on the stack are succeeding.  In other
words, \lstinline{push} and \lstinline{tryPop} are formally lock
  free\elide{~\cite{?}}.

To successfully \lstinline{cas}, a core must acquire the relevant cache line in
exclusive mode\elide{~\cite{?}}, or something conceptually equivalent.  When
several cores attempt to \lstinline{cas} a common reference, they must
coordinate at the hardware level, using precious memory bandwidth and wasting
processor cycles.  A failed \lstinline{cas} is evidence that there is
\emph{contention} over a common location.  To increase the \lstinline{cas}
success rate and conserve memory bandwidth, fine-grained concurrent algorithms
employ a \emph{backoff} scheme, which here we have abstracted into a lightweight
\lstinline{Backoff} class.  The class encapsulates the simplest scheme:
busy-wait for a random amount of time that grows exponentially each time
\lstinline{once} is invoked.

@@

Both \lstinline{push} and \lstinline{tryPop} are \emph{total} operations: they
can in principle succeed no patter what state the stack is in, and fail (and
retry) only due to active interference from concurrent threads.  A true
\lstinline{pop} operation, on the other hand, is \emph{partial}: it is undefined
when the stack is empty.  Often this is taken to mean that the operation should
block until another thread moves the stack into a state on which the operation
is defined.  Partial operations introduce considerable complexity, because
\emph{all} of the operations on the data structure must potentially signal
blocked threads, depending on the changes being performed.  In some cases, it is
possible to cleverly treat signalling in essentially the same way as atomic
updates~\cite{Scherer2004}.

Synchronization and signalling is also subject to cache and memory bandwidth
concerns, but it would take us too far afield to discuss these in depth.
Mellor-Crummey and Scott pioneered the now-common techniques for grappling with
these concerns~\cite{Mellor-Crummey1991}, and we apply their techniques in
implementing the blocking and signaling protocols for reagents.

\elide{
For example, when a lock is released, how much memory traffic is
generated?  If the lock is implemented as a single bit which threads attempt to
\lstinline{cas} in a tight loop, any change to that bit will result in a cascade
of memory traffic---even though only one waiting thread can actually win and
acquire the lock.  The most important strategy for cutting down this traffic was
introduced by Mellor-Crummey and Scott~\cite{?}, who suggested that waiting
threads should place themselves in a lock-free queue, and the release operation
should  % TODO: room for this??
}

\elide{
The goal of reagents is to systematize the above techniques and others into a
uniform set of combinators that make algorithms easier to express, and allows
algorithms to be composed and tailored by users.  In the next section, we will
see how to write all of the above examples, concisely and composably.
}

@@

\begin{figure}\elide{
\begin{lstlisting}[mathescape=true,frame=single]
// combinators for shared state
upd:  Ref[A] => PartialFunction[(A,B), (A,C)] 
             => Reagent[B,C]
read: Ref[A] => Reagent[Unit, A]
cas:  Ref[A] * A * A => Reagent[Unit, Unit]

@@

// combinators for computation
ret: A => Reagent[Unit, A]
computed: PartialFunction[A, Reagent[Unit, B]]
          => Reagent[A,B]
postCommit: (A => Unit) => Reagent[A,A]
\end{lstlisting}}
\begin{lstlisting}[mathescape=true,frame=single]
upd:  Ref[A] => PartialFunction[(A,B), (A,C)] 
             => Reagent[B,C]
read: Ref[A] => Reagent[Unit, A]
cas:  Ref[A] * A * A => Reagent[Unit, Unit]
swap: Endpoint[A,B] => Reagent[A,B]
+:    Reagent[A,B] * Reagent[A,B] => Reagent[A,B]
>>:    Reagent[A,B] * Reagent[B,C] => Reagent[A,C]
|||:    Reagent[A,B] * Reagent[A,C] => Reagent[A,(B,C)]  
ret:  A => Reagent[Unit, A]
computed: 
      PartialFunction[A,Reagent[Unit,B]] => Reagent[A,B]
postCommit: (A => Unit) => Reagent[A,A]
\end{lstlisting}
\nocaptionrule
\caption{The reagent combinators}
\label{fig:combinators}
\end{figure}

\elide{
// reagent invocation
r ! a       // as a reactant, r: Reagent[A,B], a: A
dissolve(r) // as a catalyst, r: Reagent[Unit, Unit]
}

Reagents are a new instance of an old idea: representing computations as data.
The computations being represented are fine-grained concurrent operations, so
a value of type \reagent{A}{B} represents a function from \lstinline{A} to
\lstinline{B} that internally interacts with a concurrent data structure
through mutation, synchronization, or both.
Unlike $\lambda$-abstractions, which encompass arbitrary code and treat that
code opaquely, reagents are built up from a limited set of \emph{combinators}
that are known to the implementation.  Existing reagents---for example, those
built by a concurrency expert---can be composed or tailored by library users
by applying additional combinators, without knowledge of the internal
implementation.

In this section, we introduce the reagent combinators and use them to build a
series of increasingly complex fine-grained concurrent algorithms.  By the end
of the section, we will have seen how to implement all of the algorithms
described in~\secref{background}, and several more besides.

\elide{
At any point a user can invoke a reagent, either as a
\emph{reactant} or a \emph{catalyst}---a distinction explained in
\secref{easy-examples}.

Reagents are patterned after Concurrent ML~\cite{?} and Haskell's STM
monad~\cite{?}, which represent (respectively) synchronous and atomic
operations as first-class values; we discuss the relationship in
\secref{related}.
%
Our contribution is giving a set of combinators appropriate for expressing and
composing fine-grained concurrent algorithms (\secref{easy-examples}) with a
clear semantics (\secref{semantics}), ... %TODO
}

@@

Memory is shared between reagents using the type \lstinline{Ref[A]} of
atomically-updatable references.\elide{\footnote{ These references are only
  accessible through reagent combinators, which allows them to provide
  functionality beyond the built-in \lstinline{AtomicReference[A]} type, such
  as the blocking logic we will see shortly.  }}
The \lstinline{upd} combinator~(\figref{combinators}) represents simple atomic
updates to references.  It takes an \emph{update function}, which tells how to
transform a snapshot of the reference cell and some input into an updated
value for the cell and some output.  Using \lstinline{upd}, we can rewrite
\lstinline{TreiberStack} in a more readable and concise way:
\begin{lstlisting}
class TreiberStack[A] {
  private val head = new Ref[List[A]](Nil)
  val push: Reagent[A, Unit] = upd(head) {
    case (xs, x) => (x:xs, ())
  }
  val tryPop: Reagent[Unit, Option[A]] = upd(head) {
    case (x::xs, ()) => (xs, Some(x))
    case (Nil,   ()) => (Nil, None)
  }
}
\end{lstlisting}
In Scala, anonymous partial functions are written as a series of cases
enclosed in braces.  For \lstinline{push} and \lstinline{tryPop}, the case
analyses are exhaustive, so the update functions are in fact total.
\lstinline{Unit} is akin to \lstinline{void}: it is a type with a single
member, written \lstinline{()}.

Being reagents, \lstinline{push} and \lstinline{tryPop} are inert values.  They
can be invoked using the \lstinline{!} method provided by reagents, which is
read ``react''.  For a \lstinline{Reagent[A,B]} the \lstinline{!} method takes
an \lstinline{A} and returns a \lstinline{B}.  Scala permits infix notation for
methods, so we can use a \lstinline{TreiberStack} \lstinline{s} by writing
\lstinline{s.push ! 42}.  The key point is that when we invoke these reagents,
we are executing \emph{exactly} the same algorithms written by hand in
\secref{background}, including the retry loop with exponential backoff; reagents
systematize and internalize common patterns of fine-grained concurrency.  By
exposing \lstinline{push} and \lstinline{tryPop} as reagents rather than
methods, we allow further composition and tailoring by a user of the data
structure (\secref{choice}, \secref{conjunction}).

While \lstinline{tryPop} handles both empty and nonempty stacks, we
can write a variant that drops the empty case:
\begin{lstlisting}
val pop: Reagent[Unit, A] = upd(head) {
  case (x::xs, ()) => (xs, x)
}
\end{lstlisting}
Now our update function is partial.  An invocation \lstinline{s.pop ! ()} will
block the calling thread unless or until the stack is nonempty.  Blocking and
signalling are entirely handled by the reagent implementation; there is
therefore no risk of a lost wakeup.  %TODO: add more here?

@@

Reagents interact through \emph{synchronous swap channels}, which consist of two
complementary \emph{endpoints}.  The method \lstinline{mkChan[A,B]} returns a pair
of type \lstinline{(Endpoint[A,B], Endpoint[B,A])}.
The combinator for communication is \lstinline{swap} (see
\figref{combinators}), which lifts an \lstinline{Endpoint[A,B]} to a
\lstinline{Reagent[A,B]}.  When two reagents communicate on opposite
endpoints, they provide messages of complementary type (\lstinline{A}
and \lstinline{B}, for example) and receive each other's value.  We
call a successful communication a \emph{reaction} between reagents.
On the other hand, if no complementary message is available,
\lstinline{swap} will block until a reaction can take place.  Once
again, the waiting and signalling protocol is managed internally by
the reagent implementation, so lost wakeups are impossible.

@@

If \lstinline{r} and \lstinline{s} are two reagents of the same type, their
\emph{choice} \lstinline{r + s} will behave like one of them,
nondeterministically, when invoked.  The most straightforward use of choice is
in the spirit of Unix's \lstinline{select}\elide{~\cite{?}}: waiting on several
signals simultaneously, while consuming only one of them.  For example, if
\lstinline{c} and \lstinline{d} are endpoints of the same type,
\lstinline{swap(c) + swap(d)} is a reagent that will accept exactly one message,
either from \lstinline{c} or from \lstinline{d}.  If neither endpoint has a
message available, the reagent will block until one of them does.  As usual, the
blocking protocol is handled by the reagent library.

A more novel use of choice is adding backoff strategies (of the kind described
in \secref{background}).  For example, we can build an elimination backoff stack
as follows:
\begin{lstlisting}
class EliminationStack[A] {
  private val s = new TreiberStack[A]
  private val (elimPop, elimPush) = mkChan[Unit,A]
  val push: Reagent[A,Unit] = s.push + swap(elimPush)
  val pop:  Reagent[Unit,A] = s.pop  + swap(elimPop)
}
\end{lstlisting}
Choice is left-biased,\elide{\footnote{The left bias makes choice more
  similar to \lstinline{orElse} in Haskell's STM than to CML's choice,
  but there are differences from both.  See \secref{related}.  }} so
when \lstinline{push} is invoked, it will first attempt to push onto
the underlying Treiber stack.  If the underlying push fails (due to a
lost CAS race), \lstinline{push} will attempt to send a message along
\lstinline{elimPush}, \emph{i.e.}, to synchronize with a concurrent
popper.  If it succeeds, the \lstinline{push} reagent completes
without ever having modified its underlying stack.

@@

The double-pop reagent implicitly gives a solution to the Dining Philosophers
problem\elide{~\cite{?}}: the two resources are atomically acquired.
%% Internally, the reagent library ensures dead- and live-lock freedom
%% for conjoined reagents.

@@

In chemistry, a reagent is a participant in a reaction; reagents are
subdivided into \emph{reactants}, which are consumed during reaction, and
\emph{catalysts}, which enable reactions but are not consumed by them.

The \lstinline{!} operator invokes a reagent as a reactant: the invocation
lasts for a single reaction, and any messages sent by the reagent are
consumed.  But sometimes it is useful for reagents to persist beyond a single
reaction, \emph{i.e.}, to act as catalysts.  For example, the following
function creates a catalyst that merges input from two endpoints and sends the
resulting pairs to another endpoint:
\begin{lstlisting}
def zip(in1: Endpoint[Unit,  A], in2: Endpoint[Unit,  B], 
        out: Endpoint[(A,B), Unit]) = 
  dissolve((swap(in1) ||| swap(in2)) >> swap(out))
\end{lstlisting}
The \lstinline{dissolve} function takes a \lstinline{Reagent[Unit, Unit]} and
introduces it as a catalyst.\footnote{For simplicity, we have not given a way
  to cancel catalysts after they have been introduced, but cancellation is
  easy to add.}  Operationally, in this example, that just means sending
messages along \lstinline{in1} and \lstinline{in2} that are marked as
catalyzing messages, and hence are not consumed during reaction.  The upshot
is that senders along \lstinline{in1} will see the catalyzing messages, look
for messages along \lstinline{in2} to pair with, and ultimately send messages
along \lstinline{out} (and similarly in the other order).

@@

\elide{
class TreeBarrier {
  public readonly Synchronous.Channel[] Arrive;
  private readonly Join j; // create j, init chans ...
  public TreeBarrier(int n) {Wire(0, n-1, () => {});}
  private void Wire(int low, int high, Action Done) {
    if (low == high) j.When(Arrive[low]).Do(Done);
    else if (low + 1 == high)
      j.When(Arrive[low]).And(Arrive[high]).Do(Done);
    else { // low + 1 < high
      Synchronous.Channel Left, Right; // init chans       
      j.When(Left).And(Right).Do(Done);
      int mid = (low + high) / 2;
      Wire(low, mid, () => Left());
      Wire(mid + 1, high, () => Right()); 
    }
  }
}
}

\subsection{Post-commit actions}
\label{sec:postcommit}

It is often desirable for a reagent to include ``post commit actions'': code to
be run after a reaction has successfully taken place, \emph{e.g.} for signaling
or spawning another thread after an operation completes.  The
\lstinline{postCommit} combinator~(\figref{combinators}) takes a function from
\lstinline{A} to \lstinline{Unit} and produces a \lstinline{Reagent[A,A]}.  The
idea is that the post-commit action will be recorded along with the input of
type \lstinline{A}, which is passed along unchanged.  Once the reaction
completes, the action will be invoked on the stored input.  The combinator is
meant to be used in sequence with other combinators that will produce the
appropriate input.  For example, the reagent
\lstinline{pop >> postCommit(println)} will print the popped element from a stack,
as soon as the pop has completed.

\subsection{Low-level operations on shared state: \lstinline{read} and \lstinline{cas}}
\label{sec:lowlevel}

Although the \lstinline{upd} combinator is convenient, it is sometimes necessary
to work with shared state with a greater degree of control.  To this end, we
include two combinators, \lstinline{read} and
\lstinline{cas}~(see~\figref{combinators}), for working directly on
\lstinline{Ref} values.  Together with the \lstinline{computed} combinator
described in~\secref{dynamic}, \lstinline{read} and \lstinline{cas} suffice to
\emph{build} \lstinline{upd}.

The \lstinline{read} combinator is straightforward: if \lstinline{r}
has type \lstinline{Ref[A]}, then \lstinline{read(r)} has type
\lstinline{Reagent[Unit, A]} and, when invoked, returns a snapshot of
\lstinline{r}.
%
The \lstinline{cas} combinator takes a \lstinline{Ref[A]} and two \lstinline{A}
arguments, giving the expected and updated values, respectively.  Unlike its
counterpart for \lstinline{AtomicReference}, a \lstinline{cas} reagent does
\emph{not} yield a boolean result.  Instead, the reagent library builds in
handling of failed CASes, including an optimistic retry loop and exponential
backoff.  A failure to CAS is, as discussed in \secref{choice}, a transient
failure that causes the reagent to retry.

\subsection{Constant and tentative reagents}
\label{sec:failproof}

% TODO: introduce these better

The \lstinline{ret} combinator~(\figref{combinators}) always succeeds,
immediately returning the given value.

Becasue choice is left-biased, it can be used together with the remaining
combinators to express \emph{tentative} reagents: if \lstinline{r} is a
\lstinline{Reagent[A,B]} then \lstinline{r?} is a
\lstinline{Reagent[A,Option[B]]} that first tries \lstinline{r} (wrapping its
output with \lstinline{Some}) and then tries \lstinline{ret(None)}, which always
succeeds.  This allows a reaction to be attempted, without retrying it when it
fails.  The \lstinline{?!} method combines \lstinline{?} and \lstinline{!} into
a single operation.

\subsection{Dynamic reagents}
\label{sec:dynamic}

@@

\begin{figure}
\begin{lstlisting}[frame=single]
class MSQueue[A] {
  private case class Node(data: A, next: Ref[Node])
  private val head = new Ref(new Node(null))  // sentinel
  private val tail = new Ref(read(head) ! ()) // sentinel
  val tryDeq: Reagent[Unit, Option[A]] = upd(head) {
    case (Node(_, Ref(n@Node(x, _))), ()) => (n, Some(x))
    case (emp,                        ()) => (emp, None)
  }
  private def findAndEnq(n: Node): Reagent[Unit,Unit] = 
    read(tail) ! () match {
      case ov@Node(_, r@Ref(null)) => // found true tail
        cas(r, null, n) >>
        postCommit { cas(tail, ov, n)? }
      case ov@Node(_, Ref(nv)) => // not the true tail
        cas(tail, ov, nv) ?! ();  // catch up tail ref
        findAndEnq(n)
    }
  val enq: Reagent[A, Unit] = computed { 
    (x: A) => findAndEnq(new Node(x, new Ref(null)))
  }
}
\end{lstlisting}
\nocaptionrule
\caption{The Michael-Scott queue, using reagents}
\label{fig:msqueue}
\end{figure}

@@

Here is a brief overview of the Michael-Scott queue.  The queue is represented
as a \emph{mutable} linked list, with a sentinel node at the head (front) of the
queue.  The head pointer always points to the current sentinel node; nodes are
dequeues by a CAS to this pointer, just like Treiber stacks (but lagged by one
node).  The tail of the queue is the unique node, reachable from the head
pointer, with a null \lstinline{next} pointer; thanks to the sentinel, such a
node is guaranteed to exist.  If the queue is empty, the same node will be the
head (sentinel) and tail.  Finally, as an optimization for enqueing, a tail
pointer is maintained with the invariant that the real tail is always reachable
from it.  The tail pointer may lag behind the true tail node, however, which
allows the algorithm to work using only single-word CAS instructions.

@@

As mentioned in~\secref{offers}, the initial continuation for reagents is the
\lstinline{Commit} continuation, shown in~\figref{impl}.  The
\lstinline{tryReact} method of \lstinline{Commit} makes the transition from
building up a \lstinline{Reaction} object to actually committing it.

@@

%\subsection{Methodology and experimental setup}
